---
title: "Class 5: Data visualization with GGPlot"
author: "Michelle Woo"
date: 2023-04-19
format: gfm
editor: visual 
---

# Base R plotting

We are going to start by generating the plot of class 04. This code is plotting the **cars** `dataset: plot(cars)`

```{r}
plot(cars)

```

# Ggplot2

First, we need to install the package (this is a one time thing!)

```{r}
# install.packages('ggplot2')
```

After that, we need to load the package

```{r}
library(ggplot2) 
```

Using the **cars** dataframe, we can generate a scatterplot using ggplot2

```{r}
ggplot(data = cars)

# add the aesthetics (labeling the x and y axis)
ggplot(data = cars) + aes(x=speed, y=dist)

# now add the geometry points to generate points in the graph, making sure to leave the () empty 
ggplot(data = cars) + aes(x=speed, y=dist) + geom_point()

# add a trendline
ggplot(data = cars) + aes(x=speed, y=dist) + geom_point() + geom_smooth()

# using geom_smooth(method = 'lm'), this better fits the data
ggplot(data = cars) + aes(x=speed, y=dist) + geom_point() + geom_smooth(method = 'lm')

# polish up the graph by adding labels and changing the theme
ggplot(data = cars) + aes(x=speed, y=dist) + geom_point() + geom_smooth(method = 'lm') + labs(title = 'Distance vs. Speed', subtitle = 'Using dataframe of cars', caption = 'BIMM143 Lab05', x = 'Speed (MPH)', y = 'Distance (Feet)') + theme_bw()


```

**Q1.** For which phases is data visualization important in our scientific workflows?

**Communication of Results, Exploratory Data Analysis, Detection of Outliers**

**All of the above**

**Q2.** True or False? The ggplot2 package comes already installed with R?

**False, using the install.packages command, ggplot2 will need to be installed in R**

**Q3.** Which plot types are typically NOT used to compare distributions of numeric variables?

**Network graphs**

**Q4.** Which statement about data visualization with ggplot2 is incorrect?

**ggplot2 is the only way to create plots in R**

**Q5.** Which geometric layer should be used to create scatter plots in ggplot2?

**geom_point()**

# Plotting gene expression data

Loading the data from the URL

```{r}

url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Generate ggplot

```{r}

#an initial ggplot
ggplot(data = genes) + aes(x=Condition1, y=Condition2) + geom_point()

```

**Q6.** Use the `nrow()` function to find out how many genes are in this dataset. What is your answer **5196**

**Q7.** Use the `colnames()` function and the `ncol()` function on the `genes` data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find? **4**

**Q8.** Use the `table()` function on the `State` column of this data.frame to find out how many 'up' regulated genes there are. What is your answer? **127 up regulated genes**

**Q9.** Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset? **2.44**

Figuring out the amounts of data in the gene expression data

```{r}
nrow(genes)
ncol(genes)
colnames(genes)

table(genes[,'State'])

```

Adding color to the plot

```{r}
p1 <- ggplot(data = genes) + aes(x=Condition1, y=Condition2, col=State) + geom_point()
p1

```

Changing the color scheme, adding some labels

```{r}
p2 <- p1 + scale_color_manual(values = c('blue', 'gray', 'red'))
p2

# adding labels 
p2 + labs(title = 'Differential Gene Expression', x = 'Control (no drug)', y = 'Drug Treament', caption = 'BIMM143 Class05')


```
